Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance

### 2019-04-08 Manuscript/manuscript.Rmd
manuscript edits

We first evaluated performances for models implemented using OC predictor types, without re-sampling (on the GM12878 cell line), as these characteristics represented models with the least granularity and could act as a baseline. In evaluating such composite metrics like F1-Score and MCC, we found that these models exhibited poor performances across resolutions (Figure 5A). Furthermore, at lower resolutions (50 kb and 100 kb specifically), the F1-Score and MCC could not be determined given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives <!--, as can be seen by the missing red points-->. When including OP perdictor types in the models, coupled with no resampling, we found that performances were worse than when using OC predictor types (Figure 5B). As with OC predictor types, these models were also not able to yield F1-Scores and MCCs at the 50 kb and 100 kb resolution. Models built on distance-type predictors, however, exhibited slightly better performances than the other two types, yet still yielded missing or low F1-Scores and MCC at lower resolutions (Figure 5C). Similar results were obtained on the K562 cell line, in which F1-Scores and MCCs were either low or missing across resolutions, with marginal improvement when considering distance types (Supplementary Figure 10 A-C). This indicated that the key driver leading to poor performances was not due to feature engineering, but instead the severity in the class imbalance.

@@

Comparisons of the effects of the various resampling techniques were then made. On the GM12878 cell line, we found that there was a substantial increase in F1-Score and MCC for models built on resampled data using any of the 3 techniques (Figure 5). After resampling, we were able to report performances at the lower resolutions which was not feasible for models built on fully imbalanced data. For all predictor types, implementation of either RUS or SMOTE resampling was found to yield consistently better results than no resampling and ROS. These results were confirmed when evaluating the K562 cell line (Supplementary Figure 10 A-C).

## USING DISTANCE-TYPE PREDICTORS FURTHER IMPROVED MODEL PERFORMANCES

On the GM12878 cell line, results showed that models with distance-type predictors outperformed models with other predictor types (Figure 5C). This result was consistent across both resampling technique and resolution. Directly comparing model performances using distance type predictors, we found that SMOTE resampled datasets yielded the highest model performances across each of the composite metrics. Although RUS performed only marginally worse than SMOTE, there was a steap drop off in performances for ROS and no re-sampling, across resolution. The same combination of resampling technique and predictor type was also found to result in the best performing model for the K562 cell line (Supplementary Figure 10C). These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that incorporates both random under- and over-sampling, as SMOTE is known to do.

Furthermore, we found that when using distance-type predictors, model performances tended to decline as data resolution decreased. There was a slight improvement in F1-Score for the 25 kb resolution on the GM12878 cell line, followed by a steep decline at 50 kb. Both F1-Scores and MCCs showed a momentary increase in performance at the lowest resolution considered in the study. Performances from the K562 cell line degraded much faster as resolution decreased, while the worst performing models for the GM12878 cell line were at the 50 kb resolution, followed by a slight improvement in performance at the 100 kb resolution.

>>>>

We first evaluated performances for models implemented using OC predictor types, without re-sampling, as these characteristics represented models with the least granularity and could act as a baseline. In evaluating such composite metrics like F1-Score and MCC, we found that these models exhibited poor performances across resolutions, for both cell lines (Figure 5A). Furthermore, at some resolutions , the F1-Score and MCC could not be determined given by the model's complete inability to correctly classify any genomic bins that overlapped with flanked TAD boundaries in the test data, thereby resulting in 0 true positives <!--, as can be seen by the missing red points-->. When including OP perdictor types in the models, coupled with no resampling, we found that performances were worse than when using OC predictor types (Figure 5B). As with OC predictor types, these models were also not able to yield F1-Scores and MCCs at the 50 kb and 100 kb resolution for the GM12878 cell line or any resolution lower than 10 kb for the K562 cell line. Models built on distance-type predictors, however, exhibited slightly better performances than the other two types, yet still yielded missing or low F1-Scores and MCC at lower resolutions (Figure 5C). Models built on 100 kb resolutions were not able to yeild F1-Scores for either cell line. This indicated that the key driver leading to poor performances was not due to feature engineering, but instead the severity in the class imbalance.
<!--Mention AUC is not good metric using supplementary table-->

@@

Comparisons of the effects of the various resampling techniques were then made. When models were built on randomly oversampled data, we found a substantial increase in F1-Score and MCC for both cell lines (Figure 5 D-F). The biggest improvement in performance was seen at the 10 kb resolution, with a sharp downward trend as resolutions decreased. Although performances were low at lower resolutions, both F1-Scores and MCC values were able to be obtained, contrary to models built on completely unbalanced data as seen before. Distance-type predictors were found to offer the highest F1-Scores and MCCs for randomly oversampled data, across all resolutions and for both cell lines.

Further improvement in performance was found when models were implemented on both randomly undersampled (Figure 5 G-H) and SMOTE resampled data (Figure 5 J-K). Although the differences in performance between the two techniques were marginal, SMOTE resampled data was found to yield the best F1-Scores and MCCs among all classes of models develeped and compared in the ensemble framework. These results imply that a best performing model, one that is able to best discriminate between genomic bins that are near versus far away from TAD boundaries, is one that incorporates both random under- and over-sampling, as SMOTE is known to do.

## DISTANCE-TYPE PREDICTORS PROVIDED THE MOST INFORMATIVE SPATIAL REPRESENTATION OF EPIGENOMIC ANNOTATIONS ASSOCIATED WITH TAD BOUNDARIES

Regardless, of resampling technique however, the inclusion of distance type predictors resulted in better F1-Scores and MCCs when compared to either OC or OP predictor types. This was consistent across all resolutions, for both cell lines. Furthermore, in the best overall model (SMOTE with distance-type predictors) we found that the degradation in model performance as resolution decreased was much more pronounced in the K562 cell line.  on the GM12878 cell line, there was a slight improvement in F1-Score for the 25 kb resolution, followed by a steep decline at 50 kb. Then, both F1-Scores and MCCs showed a momentary increase in performance at the lowest resolution considered in the study. Performances from the K562 cell line degraded much faster as resolution decreased, the worse being at the 100 kb resolution.
